Automatically Proving Equivalence by Type-Safe Reflection

### 2017-03-29 paper-CICM-17/content/3-hierarchyOfTactics.tex
CICM 17 : biblio to continue

The hierarchy of interfaces continues with Commutative-Monoid, Group, Commutative Group, Semi-Ring, Ring, but we only show Group for reasons of space :

@@

When trying to prove an equivalence $x \simeq y$, the universally-quantified variables have been abstracted, and they are now part of the context. Our tatics will work by normalising both sides of the "potential equivalence" $x \simeq y$, and afterwards comparing them with a syntactic equality test. The problem is that such a normalisation function will need to do different treatments for a a variable, for a constant and for an operator. For this reason, we will need to work by reflection. There will be one type of reflected terms for each algebraic structure. The novelty is not the use of reflected terms (this has already been done, as discussed in~\ref{references}), but it is the use of a type-safe reflection mechanism where we index the reflected terms by the concrete value that they represent. Each of these datatype is parametrised over a type $c$, which is the real type on which we want to prove equalities (the $carrier$ type). It is also indexed over an implementation of the corresponding interface for $c$ (we usually call it $p$, because it behaves as a $proof$ telling that the structure $c$ has the desired properties), indexed over a context of abstracted variables (a vector $\Gamma$ of $n$ elements of type $c$), and most important it is also indexed over a value of type $c$, which is precisely the concrete value being encoded.

>>>>

The hierarchy of interfaces continues with Commutative-Monoid, Group, Commutative Group, Semi-Ring and Ring, but we only show the Group interface for reasons of space :

@@

When trying to prove an equivalence $x \simeq y$, the universally-quantified variables have been abstracted, and they are now part of the context. Our tatics will work by normalising both sides of the "potential equivalence" $x \simeq y$, and afterwards comparing them with a syntactic equality test. The problem is that such a normalisation function will need to do different treatments for a a variable, for a constant and for an operator. For this reason, we will need to work by reflection. There will be one type of reflected terms for each algebraic structure. The novelty is not the use of reflected terms (this has already been done, as discussed in section~\ref{sect:relatedWork}), but is the use of a type-safe reflection mechanism where we index the reflected terms by the concrete value that they represent. Each of these datatype is parametrised over a type $c$, which is the real type on which we want to prove equalities (the $carrier$ type). It is also indexed over an implementation of the corresponding interface for $c$ (we usually call it $p$, because it behaves as a $proof$ telling that the structure $c$ has the desired properties), indexed over a context of abstracted variables (a vector $\Gamma$ of $n$ elements of type $c$), and most important it is also indexed over a value of type $c$, which is precisely the concrete value being encoded.

### 2017-03-29 paper-CICM-17/content/4-relatedWork.tex
CICM 17 : biblio to continue

As mentioned, Coq's implementation of a ring prover was also using proof-by-reflection techniques, but without the guarantees obtained with our type-safe reflection mechanism. Their automatic reflection is programmed in Ltac~\cite{DelahayeLTac} : a proof dedicated and untyped meta-language for the writing of automations. \\

@@

$Mtac$~\cite{Ziliani13} is an extension to Coq that supports dependently-typed tactic programming that uses monad to avoid the need to touch the trusted kernel typechecker of Coq. \\

Finally, when we leave the ground of nice mathematical structures like groups, rings and fields, one can decide to work with arbitrary rewriting rules, but in the general case there isn't a complete decision procedure for such problems, because there's usually no normal form. This is where deduction modulo~\cite{Dowek03} ~\cite{DelahayeModulo} and heuristics for proof search starts.

>>>>

As mentioned, Coq's implementation of a ring prover was also using proof-by-reflection techniques, but without the guarantees obtained with our type-safe reflection mechanism, and without the correct-by-construction approach : they have first defined the normalisation of terms, and afterwards they've proven the correctness of the normalisation with the external lemma :
$\forall\ (e1\ e2\ :\ Expr),\ beq_{Expr}\ (norm\ e1)\ (norm\ e2)\ =\ true\ \rightarrow\ reify\ e1\ \simeq\ reify\ e2$ which therefore needs a $reify$ function that we do not need in our development.
Our automatic reflection was written with Idris reflection mechanism which allows to do pattern matching on ssyntax, and their automatic reflection is programmed in Ltac~\cite{DelahayeLTac} : a proof dedicated and untyped meta-language for the writing of automations. \\

@@

Finally, when we leave the ground of nice mathematical structures like groups, rings and fields, one can decide to work with arbitrary rewriting rules, but in the general case there isn't a complete decision procedure for such problems, because there's usually no normal form. This is where deduction modulo~\cite{Dowek03} ~\cite{DelahayeModulo} and heuristics for proof search starts. \\

We also need to mention $Mtac$~\cite{Ziliani13}, which is an extension to Coq that supports dependently-typed tactic programming that uses monad to avoid the need to touch the trusted kernel typechecker of Coq.
